---
title: "cce_seabird_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cce_seabird_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(alohakez)
library(tidyverse)
```

### take a look at data first
### convert data types
```{r}
head(cce_seabird)
tail(cce_seabird)
cce_seabird$year <- as.integer(cce_seabird$year)
unique(cce_seabird$season)
cce_seabird$season <- factor(cce_seabird$season, 
                                levels = c("fall", "wint", "sprg", "summ"))
cce_seabird %>% head()
```

### how many years of data doo we have? do we have data for all 4 seasons?
```{r}
cce_seabird %>% group_by(year) %>% count()
cce_seabird %>% group_by(season) %>% count()
```

## Calculate Annual Average Bird Density and Richness
```{r}
annual_seabird <-  cce_seabird %>%
  group_by(year) %>%
  summarise(avg_density=mean(bird_density),
            avg_richness=mean(bird_richness))
```

```{r}
cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_density, color=season), size=1) +
  geom_line(aes(y=avg_density), data=annual_seabird, color="black", size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Density",
        title = "Seasonal Seabird Density in 1987 - 2006")

cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_richness, color=season), size=1) +
  geom_line(aes(y=avg_richness), data=annual_seabird, size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Richness",
        title = "Seasonal Seabird Richness in 1987 - 2006")
```

#### density is decreasing for all four seasons, and annually mean

#### seems like richness has been decreasing till 1997 and then going back

#### might be result of missing data for some seaons in some year

### too many lines, let's see each season separately
```{r}
cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_density, color=season), size=1) +
  geom_line(aes(y=avg_density), data=annual_seabird, color="black", size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Density",
        title = "Seasonal Seabird Density in 1987 - 2006") +
  facet_wrap(~season)

cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_richness, color=season), size=1) +
  geom_line(aes(y=avg_richness), data=annual_seabird, size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Richness",
        title = "Seasonal Seabird Richness in 1987 - 2006") +
  facet_wrap(~season)
```
#### spring seems always above average

### let's see the average based on season
```{r}
cce_seabird %>%
  group_by(season) %>%
  summarise(avg_density=mean(bird_density),
            avg_richness=mean(bird_richness))
```

### calc difference
```{r}
seabird <- left_join(cce_seabird, annual_seabird)
head(seabird)
seabird <- seabird %>%
  mutate(diff_density=bird_density-avg_density,
         diff_richness=bird_richness-avg_richness)
head(seabird)
```

###plot by season
```{r}
seabird %>%
  ggplot(aes(x=year)) +
  geom_point(aes(y=diff_density, color=season)) +
  geom_hline(yintercept=0, linetype="longdash") +
  labs( x= "Year") +
  facet_wrap(~season)

seabird %>%
  ggplot(aes(x=year)) +
  geom_point(aes(y=diff_richness, color=season)) +
  geom_hline(yintercept=0, linetype="longdash") +
  labs( x= "Year") +
  facet_wrap(~season)
```
spring seems to have higher bird density and richness


### is there a relationship between density and richness

```{r}
cce_seabird %>%
  ggplot(aes(x=bird_density, y=bird_richness)) +
  geom_point(aes(color=season)) +
  geom_point(aes(x=avg_density, y=avg_richness), alpha=0.6, shape=15, data=annual_seabird) +
  labs(x = "Bird Density", y = "Bird Richness",
        title = "Bird Density vs Bird Richness")

cce_seabird %>%
  ggplot(aes(x=bird_density, y=bird_richness)) +
  geom_point(aes(color=season)) +
  #geom_point(aes(x=avg_density, y=avg_richness), alpha=0.6, shape=15, data=annual_seabird) +
  labs(x = "Bird Density", y = "Bird Richness",
        title = "Bird Density vs Bird Richness") +
  facet_wrap(~season)
```
more density means more richness \
makes sense


### bring in data from sea-surface temperature and see if there is any relationship

### prep data first (Filter valid data: flag 0 stands for good data)
```{r}
cce_sst <- cce_sst %>%
  filter(sea_surface_temperature_flag==0) %>%
  filter(!is.na(sea_surface_temperature_c)) %>%
  filter(year>=1980)                            #save computing
cce_sst <- cce_sst %>%
    mutate(month_day=as.Date(paste(2224, month, day, sep = "/" )))
```

#### now create season variable (see cce_sst for explanation)
```{r}
fall_start <- as.Date("2224/09/22")
wint_start <- as.Date("2224/12/21")
sprg_start <- as.Date("2224/3/20")
summ_start <- as.Date("2224/6/20")

cce_sst <- cce_sst %>% mutate(
  season = case_when( month_day>=fall_start &  month_day<wint_start ~ "fall",
                      month_day>=wint_start | month_day<sprg_start  ~ "wint",
                      month_day>=sprg_start &  month_day<summ_start ~ "sprg",
                      month_day>=summ_start &  month_day<fall_start ~ "summ"
                     )
  ) 
cce_sst$season <- factor(cce_sst$season, 
                                levels = c("fall", "wint", "sprg", "summ"))
levels(cce_sst$season)[2:3]
#cce_sst %>% select(date_pst:day, month_day, season)
```
##### calculate seasonal temperature
```{r}
annual_sst<- cce_sst %>%
  group_by(year) %>%
  summarize(annual_temp = mean(sea_surface_temperature_c))
annual_sst %>% head()
seasonal_sst <- cce_sst %>%
  group_by(year, season) %>%
  summarize(seasonal_temp = mean(sea_surface_temperature_c))
seasonal_sst <- left_join(seasonal_sst, annual_sst)
```

##### merge data together
```{r}
bird_sst <- left_join(seabird, seasonal_sst) %>% arrange(year, season)
bird_sst  %>% select(year, season, seasonal_temp, annual_temp)
```
#### before exploring relationship, see sst in these years
```{r}
bird_sst %>%
  ggplot(aes(x=year)) +
  geom_path(aes(y=seasonal_temp, color=season), size=1) +
  geom_path(aes(y=annual_temp)) +
  theme_minimal() +
  labs(x = "Year", y = "Seasonal Sea-Surface Temperature (C)")
```

#### see shape 
```{r}
bird_sst %>%
  ggplot(aes(x=year)) +
  geom_path(aes(y=annual_temp), color="blue") +
  geom_path(aes(y=avg_density)) +
  geom_path(aes(y=avg_richness)) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10))
```
#### there was a drop in sst, density, and richness between arround 96-99


#####now 
```{r}
bird_sst %>%
  ggplot() +
  geom_point(aes(x=bird_density, y= seasonal_temp, color=season)) +
  geom_point(aes(x=avg_density, y=annual_temp), alpha=0.6, shape=15, size=1)

bird_sst %>%
  ggplot() +
  geom_point(aes(x=bird_density, y= seasonal_temp, color=season)) +
  geom_point(aes(x=avg_density, y=annual_temp), alpha=0.6, shape=15, size=1) +
  facet_wrap(~season)
```


```{r}
bird_sst %>%
  ggplot() +
  geom_point(aes(x=bird_richness, y= seasonal_temp, color=season)) +
  geom_point(aes(x=avg_richness, y=annual_temp), alpha=0.6, shape=15)
```
we see lower bird richness happens when temperature is lower than 18 
