---
title: "cce_watertemp_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cce_watertemp_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(alohakez)
library(tidyverse)
```

## Let's take a look at the data first
```{r}
cce_watertemp %>% head()
sapply(cce_watertemp, typeof)
```

## Filter valid data: only 0 stands for good date 
```{r}
watertemp <- cce_watertemp %>%
  filter(sea_surface_temperature_flag==0) %>%
  filter(!is.na(sea_surface_temperature_c))
```

## Calculate annual average 
## Calculate lowest and highest temperature over years

```{r}
annual_temp <- watertemp %>%
  group_by(year) %>%
  summarize(annual_avg = mean(sea_surface_temperature_c,),
            min_temp = min(sea_surface_temperature_c),
            max_temp = max(sea_surface_temperature_c)) %>%
  mutate(diff=max_temp-min_temp)
```

### We look at the annual temperature over years
```{r}
annual_temp %>%
  ggplot(aes(x = year, y = annual_avg)) +
  geom_point() +
  labs(x = "Year",
       y = "Mean Sea-Surface Temperature (C)",
       title = "Annual Mean Sea-Surface Temperature from 1916 to 2015 (C)")
```

### now see the min and max over years
```{r}
annual_temp %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y=min_temp), color="blue") +
  geom_point(aes(y=max_temp), color="orange") +
  labs(x = "Year",
       y = "Sea-Surface Temperature (C)",
       title = "Lowest and Highest Day Sea-Surface Temperature (C)")
```

### is there anything interesting about the range in years
```{r}
annual_temp %>%
  ggplot(aes(x = year, y = diff)) +
  geom_point() +
  labs(x = "Year",
       y = "Sea-Surface Temperature (C)",
       title = "Difference") 
```

## Calculate monthly average, minimum, and maximum
```{r}
monthly_temp <- watertemp %>%
  group_by(year, month) %>%
  summarize(monthly_avg = mean(sea_surface_temperature_c),
            min_temp = min(sea_surface_temperature_c),
            max_temp = max(sea_surface_temperature_c))
```

### let's see how temp varies over month for recent years
```{r}
monthly_temp %>%
  filter(year>2009) %>%
  ggplot() +
  geom_point(aes(x = month, y = monthly_avg)) +
  labs(x = "Month",
       y = "Mean Sea-Surface Temperature (C)",
       title = "Sea-Surface Temperature (C) in Recent Years") +
  facet_wrap(~year)
```

## when does temp reach highest 
```{r}
highest_temp <-watertemp %>%
    mutate(date_pst=as.Date(date_pst, "%Y/%m/%d")) %>%
    group_by(year) %>%
    slice(which.max(sea_surface_temperature_c))
highest_temp <- highest_temp %>%
    mutate(month_day=as.Date(paste(2222, month, day, sep = "/" )))
highest_temp %>% head()
highest_temp %>% tail()
```

### map it
```{r}
highest_temp %>%
  ggplot() +
  geom_path(aes(x=year, y=month_day)) +
  scale_y_date(date_breaks = "10 days", date_labels = "%b%d")
```

### same thing for lowest
```{r}
lowest_temp <-watertemp %>%
    mutate(date_pst=as.Date(date_pst, "%Y/%m/%d")) %>%
    group_by(year) %>%
    slice(which.min(sea_surface_temperature_c))
lowest_temp <- lowest_temp %>%
    mutate(month_day=as.Date(paste(2222, month, day, sep = "/" )))
lowest_temp %>% head()
lowest_temp %>% tail()
```

### map it
```{r}
lowest_temp %>%
  ggplot() +
  geom_point(aes(x=year, y=month_day)) +
  scale_y_date(date_breaks = "1 month", date_labels = "%b%d")
```

#### intersting, but hard to see, let's modify the month_day variable
#### we'll chose November as a cutoff
```{r}
lowest_temp <-watertemp %>%
    mutate(date_pst=as.Date(date_pst, "%Y/%m/%d")) %>%
    group_by(year) %>%
    slice(which.min(sea_surface_temperature_c))
lowest_temp <- lowest_temp %>%
    mutate( month_day = as.Date(ifelse(month>=11,
                               paste(2222, month, day, sep = "/" ), paste(2223, month, day, sep = "/" ) 
                               )) )
lowest_temp %>% head()
lowest_temp %>% tail()
```

### map again
```{r}
lowest_temp %>%
  ggplot() +
  geom_point(aes(x=year, y=month_day)) +
  scale_y_date(date_breaks = "1 month", date_labels = "%b%d")
```

