---
title: "cce_seabird_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cce_seabird_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(alohakez)
library(tidyverse)
```

### take a look at data first
### convert data types
```{r}
head(cce_seabird)
tail(cce_seabird)
cce_seabird %>% sapply(typeof)
cce_seabird$year <- as.integer(cce_seabird$year)
unique(cce_seabird$season)
cce_seabird$season <- factor(cce_seabird$season, 
                                levels = c("fall", "wint", "sprg", "summ"))
cce_seabird %>% sapply(typeof)
cce_seabird %>% head()
```

### how many years of data doo we have? do we have data for all 4 seasons?
```{r}
cce_seabird %>% group_by(year) %>% count()
cce_seabird %>% group_by(season) %>% count()
```

## Calculate Annual Average Bird Density and Richness
```{r}
annual_seabird <-  cce_seabird %>%
  group_by(year) %>%
  summarise(avg_density=mean(bird_density),
            avg_richness=mean(bird_richness))
```

```{r}
cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_density, color=season), size=1) +
  geom_line(aes(y=avg_density), data=annual_seabird, color="black", size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Density",
        title = "Seasonal Seabird Density in 1987 - 2006")

cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_richness, color=season), size=1) +
  geom_line(aes(y=avg_richness), data=annual_seabird, size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Richness",
        title = "Seasonal Seabird Richness in 1987 - 2006")
```

#### density is decreasing for all four seasons, and annually mean

#### seems like richness has been decreasing till 1997 and then going back

#### might be result of missing data for some seaons in some year

### too many lines, let's see each season separately
```{r}
cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_density, color=season), size=1) +
  geom_line(aes(y=avg_density), data=annual_seabird, color="black", size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Density",
        title = "Seasonal Seabird Density in 1987 - 2006") +
  facet_wrap(~season)

cce_seabird %>%
  ggplot(aes(x=year)) +
  geom_line(aes(y=bird_richness, color=season), size=1) +
  geom_line(aes(y=avg_richness), data=annual_seabird, size=0.8, linetype="dotdash") +
  labs( x= "Year", y = "Bird Richness",
        title = "Seasonal Seabird Richness in 1987 - 2006") +
  facet_wrap(~season)
```
#### spring seems always above average

### let's see the average based on season
```{r}
cce_seabird %>%
  group_by(season) %>%
  summarise(avg_density=mean(bird_density),
            avg_richness=mean(bird_richness))
```

### calc difference
```{r}
seabird <- left_join(cce_seabird, annual_seabird)
head(seabird)
seabird <- seabird %>%
  mutate(diff_density=bird_density-avg_density,
         diff_richness=bird_richness-avg_richness)
head(seabird)
```
```{r}
seabird %>%
  ggplot(aes(x=year)) +
  geom_point(aes(y=diff_density, color=season)) +
  geom_hline(yintercept=0, linetype="longdash") +
  labs( x= "Year") +
  facet_wrap(~season)

seabird %>%
  ggplot(aes(x=year)) +
  geom_point(aes(y=diff_richness, color=season)) +
  geom_hline(yintercept=0, linetype="longdash") +
  labs( x= "Year") +
  facet_wrap(~season)
```


### is there a relationship between density and richness

```{r}
cce_seabird %>%
  ggplot(aes(x=bird_density, y=bird_richness)) +
  geom_point(aes(color=season)) +
  geom_point(aes(x=avg_density, y=avg_richness), alpha=0.6, shape=15, data=annual_seabird) +
  labs(x = "Bird Density", y = "Bird Richness",
        title = "Bird Density vs Bird Richness")

cce_seabird %>%
  ggplot(aes(x=bird_density, y=bird_richness)) +
  geom_point(aes(color=season)) +
  #geom_point(aes(x=avg_density, y=avg_richness), alpha=0.6, shape=15, data=annual_seabird) +
  labs(x = "Bird Density", y = "Bird Richness",
        title = "Bird Density vs Bird Richness") +
  facet_wrap(~season)
```
### more density means more richness \
### makes sense


### bring in data from sea-surface temperature and see if there is any relationship

