---
title: "pie_crab: Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{arc_weather}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE, warning = FALSE
)
```

```{r setup, message = FALSE, warning = FALSE}
library(alohakez)
library(tidyverse)
library(rstatix)
library(GGally)
```

```{r}
pie_crab %>% group_by(date, latitude, site, mata, matw) %>%
  count() %>% arrange(latitude)
max(pie_crab$date) - min(pie_crab$date)
max(pie_crab$latitude) - min(pie_crab$latitude)
```
generally, higher latitude, lower air and water temp
13 sites were sampled, within 20 days, between july 24 and aug 13
each site had 25 - 37 crabs,

```{r}
pie_crab %>% group_by(latitude, site) %>%
  get_summary_stats(size, type = c("common")) %>%
  select(latitude, site, n, mean, min, max, sd)
levels(pie_crab$site)
pie_crab <- pie_crab %>%
  mutate(site = fct_reorder(site, latitude, min),
         name = fct_reorder(name, latitude, min))
```
```{r}
pie_crab %>%
  ggplot(aes(y=as.numeric(factor(latitude)), group = latitude)) +
  geom_boxplot(aes(size)) +
  scale_x_continuous(breaks = seq(from = 6, to = 24, by = 6)) +
  scale_y_continuous(
    breaks = 1:length( levels(pie_crab$site) ),
    labels = levels( factor(pie_crab$latitude) ),
    sec.axis = sec_axis(~.,
                        breaks = 1:length( levels(pie_crab$site) ),
                        labels = levels(pie_crab$name)
                     )
  ) +
  theme(axis.text.y.right = element_text(colour = "darkblue")) +
  labs(y = "latitude of salt marsh",
       x = "carapace width of crab (mm)")
```


```{r}
crab <- pie_crab %>% group_by(latitude, mata, sata, matw, satw) %>%
  summarise(mean_size = mean(size), max_size = max(size))
```


```{r}
mod0 <- lm(mean_size~ latitude, data = crab)
summary(mod0)

mod1 <- lm(mean_size~ mata+sata+matw+satw, data = crab)
anova(mod1)
summary(mod1)
vif(mod1)

mod2 <- lm(mean_size~ sata+matw+satw, data = crab)
anova(mod2)
vif(mod2)

mod3 <- lm(mean_size~ matw+satw, data = crab)
anova(mod3)
vif(mod3)
summary(mod3)
```


```{r}
mod0 <- lm(max_size~ latitude, data = crab)
summary(mod0)

mod1 <- lm(max_size~ mata+sata+matw+satw, data = crab)
summary(mod1)
vif(mod1)

mod2 <- lm(max_size~ sata+matw+satw, data = crab)
summary(mod2)
vif(mod2)

mod3 <- lm(max_size~ matw+satw, data = crab)
anova(mod3)
vif(mod3)
summary(mod3)
```


```{r}
# Citation

Johnson, D. 2019. Fiddler crab body size in salt marshes from Florida to Massachusetts, USA at PIE and VCR LTER and NOAA NERR sites during summer 2016. ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/4c27d2e778d3325d3830a5142e3839bb (Accessed 2021-05-27).

Johnson DS, Crowley C, Longmire K, Nelson J, Williams B, Wittyngham S. The fiddler crab, Minuca pugnax, follows Bergmann's rule. Ecol Evol. 2019;00:1–9. https://doi.org/10.1002/ece3.5883

# How we processed the raw data

`r knitr::spin_child(here::here("data-raw","pie_crab.R"))`
