---
title: "vcr_fish_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vcr_fish_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(alohakez)
library(tidyverse)
```

## take a look at data first
```{r}
vcr_fish %>% head()
summary(vcr_fish)
typeof(vcr_fish$sample_date)
typeof(vcr_fish$sample_time)
```

## data pre-processing
```{r}
vcr_fish <- vcr_fish %>%
  mutate( sample_date=as.Date(sample_date, "%Y-%m-%d") ) %>%
  mutate(sample_year=as.numeric( format(sample_date, format = "%Y") ),
         sample_month=as.numeric( format(sample_date, format = "%m") ),
         sample_day=as.numeric( format(sample_date, format = "%d") )
         )
vcr_fish <- vcr_fish %>%
  mutate(sample_hour=as.numeric(
              format(as.POSIXct(sample_time, format = "%H:%M"), format = "%H")
              ),
         sample_min=as.numeric(
           format(as.POSIXct(sample_time, format = "%H:%M"), format = "%M")
           )
        )
```

### check
```{r}
sapply(vcr_fish, typeof)
```

## look at fish length
```{r}
vcr_fish %>%
  ggplot() +
  geom_histogram(aes(x=length), binwidth=30)
```
#### there is one super long fish, seems to vary across spieces

# let's see how many species were measured
```{r}
vcr_fish %>%
  count(count) 
```
# filter out Na row for count & length since we focus on fish for now
```{r}
fish <- vcr_fish %>% filter(!is.na(count) & !is.na(length)) 
```

### We want to know how many of each species do we have
```{r}
species_count <- fish %>%
  group_by(species_name) %>%
  summarise(total_count=sum(count))
species_count
```
#### from the plot, we can see there are 32 species in this data

```{r}
species_count %>%
  ggplot(aes(y=species_name) ) +
  geom_bar(aes(weight = total_count))
```

```{r}
main_species_name <- species_count %>% filter(total_count>300)
main_species_name <- main_species_name %>% 
  arrange(total_count) %>% 
  pull(species_name)
main_species_name 
```
#### "Bay Anchovy"  "Silver Perch" "Silversides"  "Pinfish"  "Pipefish"

### look at these species
```{r}
main_species_r <- fish %>% filter(species_name %in% main_species_name)
main_species_r %>% 
  group_by(species_name) %>%
  summarise(total_count=sum(count))
```

```{r}
main_species_r %>% filter(count!=1) %>% arrange(desc(count)) %>%
  select(species_name, count, length)
```
## let's see a histogram based on species

```{r}
main_species_r %>% 
  ggplot() +
  geom_histogram(aes(x=length, fill=species_name), binwidth=10)
```

### we suspect there is an outlier, remove it
```{r}
pipefish_outlier <- main_species_r %>% slice_max(length) #store it in case
main_species <- main_species_r %>% filter(length!=pipefish_outlier$length)
```

## let's plot it again
```{r}
main_species %>% 
  ggplot() +
  geom_histogram(aes(x=length, fill=species_name), binwidth=10)
```

```{r}
main_species_r %>% filter(count!=1) %>% arrange(desc(count)) %>%
  select(species_name, count, length)
```

#### there are rows where the count is bigger than one, if we want to analyze the data, it shuld be taken into account that those rows are weighed more

```{r}
count2fish <- main_species %>% filter(count==2) %>% slice(rep(1:n(), each = 2-1))
count3fish <- main_species %>% filter(count==3) %>% slice(rep(1:n(), each = 3-1))
count5fish <- main_species %>% filter(count==5) %>% slice(rep(1:n(), each = 5-1))
count6fish <- main_species %>% filter(count==6) %>% slice(rep(1:n(), each = 6-1))
count19fish <- main_species %>% filter(count==19) %>% slice(rep(1:n(), each = 19-1))
count50fish <- main_species %>% filter(count==50) %>% slice(rep(1:n(), each = 50-1))
count56fish <- main_species %>% filter(count==56) %>% slice(rep(1:n(), each = 56-1))
main_species <- main_species %>%
  bind_rows(count2fish, count3fish, count5fish, count6fish,
            count19fish, count50fish, count56fish)
```


## draw desnity plot of these 3 fishes
```{r}
main_species %>% 
  ggplot() +
  geom_histogram(aes(x=length, fill=species_name), binwidth=10) +
  labs(y="fish length", 
       title="Histogram of fish length based on species")
```

```{r}
main_species %>%
  ggplot(aes(x=length)) +
  geom_histogram(aes(y = ..density.., 
                     fill=species_name, color=species_name), binwidth = 10) +
  geom_density(aes(fill=species_name), alpha = 0.8, size=0.2) +
  labs(y="fish length", 
       title="density plot of fish length")
```
```{r}
main_species %>%
  ggplot(aes(length, fill=species_name, color=species_name)) +
  geom_density(alpha = 0.6, position = "stack", binwidth = 10) + 
  labs(y="fish length", 
       title="stacked density plot of fish length")
```

### see each species
```{r fig.width=10}
main_species %>%
  ggplot(aes(length, fill=species_name)) +
  geom_histogram(binwidth=15) +
  facet_wrap(~species_name)
```

```{r fig.width=10}
main_species %>%
  ggplot(aes(length, fill = species_name, color=species_name)) +
  geom_density(alpha=2) +
  facet_wrap(~species_name)
```

### look at pipefish, intersting, why 2 clusters
```{r}
main_species %>% filter(species_name=="Pipefish") %>%
  group_by(sample_year, sample_month) %>% count()
```
```{r fig.width=15, fig.height=10}
main_species %>% filter(species_name=="Pipefish") %>%
  ggplot(aes(length, fill = species_name, color=species_name)) +
  geom_histogram(binwidth=10) +
  facet_wrap(~site)
```


```{r}
main_species %>% filter(species_name=="Pipefish") %>%
  ggplot() +
  geom_point(aes(x=sample_date, y=length, color=site)) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
  theme(axis.text.x = element_text(angle = 90, hjust=0, vjust=0)) 
```

